# Compilation of international catches

## Compile catch in CAR and NED jurisdictional waters by discards and reporting type

Now that we have separated the EEZ catch by region, we will summarize it according to the areas used in the dolphin management strategy evaluation operating model. For the Caribbean region, we have the high and low international catch scenarios that we compile for the sensitivity analyses in the operating model.  

```{r}
#| echo: true

rm(list = ls())  # clear the workspace

# load data
load("data/outputs/SAU_EEZs_WCA.RData")  # SAU EEZ data for NED and NCA saved from earlier

# summarize the catch by area and separate out Canadian landings
tapply(dwest$lbs, dwest$reg, sum, na.rm = T)

dwest$year <- factor(dwest$year, levels = 1950:2019)
dc <- dwest[which(dwest$reg == "Canadian EEZ"), ]

CAN <- tapply(dc$lbs, dc$year, sum, na.rm = T)
#tail(CAN)

# load Caribbean international landings scenarios 
car_sc <- read.csv("data/hi_lo_CAR_landings.csv")
car_sc <- car_sc[which(car_sc$year >= 1986), ]    # only need data 1986 and onwards
head(car_sc)

# combine CAR reported landings, unreported discards, unreported landings
CARlo <- car_sc$lo
CARhi <- car_sc$hi

```

## Distribute the territorial international catches across quarters of the year

The international catches are only reported as annual catches and do not have any month or season associated with them. However, the MSE operating model requires seasonal catches. Lacking other information on the seasonality of these fleets, we will assume that the fleet distribution of the U.S. pelagic longline is representative of the seasonality of the jurisdictional fleets. Now we parse the annual catches according to the seasonality of the U.S. fleet.

The pelagic longline logbook data is only available going back to 1997, but there are Caribbean catches extending back to the 1950s. To interpolate the seasonality for 1986 - 1996, we take the average seasonality of the oldest 10 years of logbook data (1997 - 2006) and use those averages for the unknown years.  

```{r}
#| echo: true

# read in the percentage of the catch by area and year-quarter combinations from the U.S. PLL
load("data/outputs/per_PLLcatch_by_area_yearquarter.RData")  

names(percatch[1, 1, ]) # matrix 2 == Caribbean, matrix 7 == NED (Atlantic Northwest)

# parse Canada data into quarters
CAN <- CAN[which(rownames(CAN) >= 1997)]  # PLL data are only available from 1997 on 
CAN <- c(CAN, CAN[length(CAN)], CAN[length(CAN)], CAN[length(CAN)])  # interpolate 2019 catches for 2020-2022

CANqrt <- percatch[, , 7]   # create matrix for final catch data
for (i in 1:ncol(percatch))  { 
  CANqrt[, i] <- percatch[, i, 7] * CAN[i]  }

percatch_car <- percatch[, , 2] # create matrix for Caribbean percentages
int <- rowMeans(percatch_car[, 1:10]) # calculate the average seasonality using the oldest 10 years of logbook data
percatch_car <- cbind(matrix(rep(int, length(1986:1996)), nrow = 4), percatch_car)  # attach those 10-year averages to years 1986:1996
colnames(percatch_car) <- 1986:2022   # full data frame for Caribbean seasonality 1986 - 2022

length(CARlo)
dim(percatch_car)

CARlo_qrt <- percatch_car   # create matrices for final catch data
CARhi_qrt <- percatch_car

for (i in 1:ncol(percatch_car))  { 
  CARlo_qrt[, i] <- percatch_car[, i] * CARlo[i]
  CARhi_qrt[, i] <- percatch_car[, i] * CARhi[i]
}

# ensure sums are equal after parsing
sum(CARlo_qrt); sum(CARlo)
sum(CARhi_qrt); sum(CARhi)
sum(CAN, na.rm = T); sum(CANqrt, na.rm = T)
```
The Canada jurisdictional data need to be combined with the NED high seas data calculated in the previous section, because both of these are a component of the international catches in the NED area. 

```{r}
#| echo: true

load("data/outputs/NED_year_quarter.RData")

NEDtot <- NED + CANqrt
# ensure sums are correct
sum(NED, na.rm = T) + sum(CANqrt, na.rm = T); sum(NEDtot, na.rm = T)

NEDtot <- NEDtot[, !is.na(colSums(NEDtot))] # remove NA columns from CAN data

# plot the parsed out data by year-quarter
yrs <- sort(rep(as.numeric(colnames(CARlo_qrt)), 4))
qrt <- rep(1:4, length(yrs)/4)
yrs1 <- sort(rep(as.numeric(colnames(NEDtot)), 4))
qrt1 <- rep(1:4, length(yrs1)/4)

matplot(yrs + qrt/4, cbind(matrix(CARhi_qrt), matrix(CARlo_qrt))/10^6, lty = 1, lwd = 2, 
        col = 2:3, type = "l", main = "International jurisdictional catches by quarter", 
        xlab = "", ylab = "total catch (millions of pounds)")
lines(yrs1 + qrt1/4, as.numeric(matrix(NEDtot))/10^6, lwd = 2, col = 4)
legend("topleft", c("Caribbean landings - high scenario", 
                    "Caribbean landings - low scenario", "NED reported landings"),        
       lwd = 2, col = 2:4, bty = "n")

```
Finally, we format the data in the format for input into the operating model. 

```{r}
#| echo: true

# format for operating model - Caribbean 
flt <- rep("Intl", length(yrs))
are <- rep("CAR", length(yrs))

# high and low landings scenarios
finhi <- data.frame(cbind(yrs, qrt, flt, are, matrix(CARhi_qrt)))
finlo <- data.frame(cbind(yrs, qrt, flt, are, matrix(CARlo_qrt)))

# format final matrices
names(finhi) <- c("Year", "Quarter", "Fleet", "Area", "Catch_lbs")
finhi$Catch_lbs <- as.numeric(finhi$Catch_lbs)
head(finhi)

names(finlo) <- c("Year", "Quarter", "Fleet", "Area", "Catch_lbs")
finlo$Catch_lbs <- as.numeric(finlo$Catch_lbs)
head(finlo)

#par(mfrow = c(2, 2))
#plot(finhi$Catch_lbs, type = "l")
#plot(finhi$Catch_lbs ~ factor(finhi$Fleet))
#plot(finhi$Catch_lbs ~ factor(finhi$Quarter))
#plot(finhi$Catch_lbs ~ factor(finhi$Year))
#apply(finhi, 2, table)

# write the output files for both scenarios
write.csv(finhi, file = "data/FINAL_files/CAR_Intl_hi_TomFormat.csv", row.names = FALSE)
write.csv(finlo, file = "data/FINAL_files/CAR_Intl_lo_TomFormat.csv", row.names = FALSE)

# format for operating model - NED 
flt <- rep("Intl", length(yrs1))
are <- rep("NED", length(yrs1))
fin <- data.frame(cbind(yrs1, qrt1, flt, are, matrix(NEDtot)), stringsAsFactors = FALSE)
names(fin) <- c("Year", "Quarter", "Fleet", "Area", "Catch_lbs")
fin$Catch_lbs <- as.numeric(fin$Catch_lbs)
head(fin)

#apply(fin, 2, table)

# write output file
write.csv(fin, file = "data/FINAL_files/NED_Intl_TomFormat.csv", row.names = FALSE)

#barplot(tapply(fin$Catch_lbs, list(fin$Quarter), sum, na.rm = T), col = rainbow(4))

```



