# NOAA Fisheries landings data

Here we will explore commercial and recreational landings data accessed through the FOSS (Fisheries One Stop Shop) data portal. The advantage of this portal is that commercial and recreational data are available side-by-side in the same units (pounds landed).

## Data access and upload

Data are downloaded from the NOAA Fisheries [FOSS site](https://www.fisheries.noaa.gov/foss/f?p=215:200:660844904940::NO:RP::). For the Data Set, we choose both commercial and recreational. For the Years, we select all years from 1980 to the present. For NMFS Region, we select the "NMFS Regions" button and then select the Gulf, South Atlantic, Middle Atlantic and New England from the box. For the Species, we select "Dolphinfish" and "Dolphinfish\*\*." For Report Format, we choose TOTALS BY YEAR/REGION. Hitting the "RUN REPORT" button then generates the data needed for the subsequent analysis, which can be downloaded as a .csv file under the "Actions" button.

We then read in the data file and clean it for analysis (removing the commas from the numbers so that R recognizes the values as numbers). We output the file header to make sure it was read in and that the numbers were converted correctly.

```{r}
#| warning: false
#| echo: true

# clear workspace
rm(list = ls())

# read in data file
d <- read.table("data/FOSS_landings.csv", header = T, sep = ",", skip = 1)
head(d)

# take commas out of file --------------------------
d$Pounds <- as.numeric(gsub(",", "", d$Pounds))
d$Dollars <- as.numeric(gsub(",", "", d$Dollars))
d$Metric.Tons <- as.numeric(gsub(",", "", d$Metric.Tons))
head(d)
```

## Summarize the landings

We then summarize the landings (in pounds) by year, sector, and region. The result is a new data frame with each row representing a year and each column representing the total landings for each region and sector.

```{r}
#| echo: true

# compile by year for sectors and regions ---------------
d$Collection <- factor(d$Collection, levels = c("Recreational", "Commercial")) 
tab <- tapply(d$Pounds, list(d$Year, d$Collection, d$Region.Name), sum, na.rm = T)
tab[is.na(tab)] <- 0
#head(tab)

dat <- data.frame(cbind(tab[, , "Gulf"], tab[, , "South Atlantic"], 
                        tab[, , "Middle Atlantic"], tab[, , "New England"]))
names(dat) <- c("rec_Gulf", "com_Gulf", "rec_SA", "com_SA", "rec_MA", "com_MA", "rec_NE", "com_NE")
dat
```

## Plot the summary

Now we plot the data, using similar color schemes by region. Because the recreational data have high imprecision prior to the 1990s we plot from 1990 to the last year of data.

```{r}
#| echo: true

# select years ---------------------
dat <- dat[which(rownames(dat) > 1990), ]

# define colors and plot -------------------
cols <- c("#0000FF30", 4, "#FF000030", 2, "#00FF0030", 3, "#00000030", 1)

par(mar = c(4, 4, 1, 0), mgp = c(2.5, 1, 0))
yrs <- as.numeric(rownames(dat))

b <- barplot(t(dat)/10^6, names.arg = yrs, col = cols, ylim = c(0, 43), 
             border = NA, las = 2, ylab = "total landings (millions of pounds)", 
             main = "Dolphin landings in the U.S. EEZ")
legend("top", c("New England - commercial", "New England - recreational", 
                "Mid-Atlantic - commercial", "Mid-Atlantic - recreational", 
                 "South Atlantic - commercial", "South Atlantic - recreational",
                 "Gulf of Mexico - commercial", "Gulf of Mexico - recreational"),
    col = cols[8:1], pch = 15, pt.cex = 1.5, bty = "n", y.intersp = 0.9, cex = 0.85)
axis(1, at = b, lab = rep("", length(b)), pos = 0)
abline(h=0)

```

## Updated Gulf/Atlantic parsing of data

There are some issues with the way that the FOSS system splits out Gulf and Atlantic catches by Monroe County, which is on the border between the two jurisdictions. We received data from Mike Larkin (SERO) which has the correct split out of data for Monroe County between the Gulf and Atlantic. We compare the total recreational catch for the Gulf and Atlantic between the FOSS data and the updated data. The numbers are very close; the average difference from the FOSS data is 2%.

```{r}
#| echo: true

ml <- read.table("data/RecreationalDolphinLandings_SAandGOM_MLarkin.csv", header = T, sep = ",", skip = 6)
ml <- ml[which(ml$Year > 1990), ]

# select same years for FOSS data set 
dat <- dat[which(yrs <= max(ml$Year)), ]

# check that years match
summary(rownames(dat) == ml$Year)
yrs <- as.numeric(rownames(dat))

# check that totals match
old <- dat$rec_Gulf + dat$rec_SA + dat$rec_MA + dat$rec_NE
new <- ml$GOM + ml$ATL

# plot totals 
par(mar = c(4, 6, 2, 1), mgp = c(2.5, 1, 0))
matplot(yrs, cbind(old, new), xlab = "", lty = 1, col = c(2, 4), type = "b", pch = 19, 
        ylab = "total recreational landings\n(Gulf + Atlantic, in pounds)",
        main = "Comparison of FOSS data with data corrected for Monroe County")
legend("topright", c("FOSS data", "corrected data"), col = c(2, 4), pch = 19, lty = 1)

# average % difference between FOSS and updated data
mean((old - new)/old*100)
# min and max % difference between FOSS and updated data
range((old - new)/old*100)
hist((old - new)/old*100, main = "Histogram of percent difference from FOSS data", breaks = 10, 
     xlab = "(FOSS - update) / FOSS")

# replace FOSS data with update data
datnew <- dat
datnew$rec_Gulf <- ml$GOM
datnew$rec_SA <- ml$ATL - (datnew$rec_MA + datnew$rec_NE)

# old data set
tail(dat)
# updated data set
tail(datnew)
```

## Plots with updated data

We replace the FOSS values for Gulf and South Atlantic recreational catches with these new values and re-plot the full data set. We can also plot the landings as a line graph so that the relative magnitudes of catch across region and sector are more easily compared.

```{r}
par(mar = c(4, 4, 1, 0), mgp = c(2.5, 1, 0))
b <- barplot(t(datnew)/10^6, names.arg = yrs, col = cols, ylim = c(0, 43), 
             border = NA, las = 2, ylab = "total landings (millions of pounds)", xlab = "",
             main = "Dolphin landings in the U.S. EEZ -- correct Monroe County split")
legend("top", c("New England - commercial", "New England - recreational", 
                "Mid-Atlantic - commercial", "Mid-Atlantic - recreational", 
                 "South Atlantic - commercial", "South Atlantic - recreational",
                 "Gulf of Mexico - commercial", "Gulf of Mexico - recreational"),
    col = cols[8:1], pch = 15, pt.cex = 1.5, bty = "n", y.intersp = 0.9, cex = 0.85)
axis(1, at = b, lab = rep("", length(b)), pos = 0)
abline(h=0)

matplot(yrs, datnew/10^6, col = c(4, 4, 2, 2, 3, 3, 1, 1), lty = c(1, 3, 1, 3, 1, 3, 1, 3), type = "l",
             las = 2, xlab = "", ylab = "total landings (millions of pounds)", lwd = 2, 
             main = "Dolphin landings in the U.S. EEZ -- correct Monroe County split")
legend("topright", c("Gulf of Mexico - recreational", "Gulf of Mexico - commercial", 
                     "South Atlantic - recreational", "South Atlantic - commercial",
                     "Mid-Atlantic - recreational", "Mid-Atlantic - commercial", 
                     "New England - recreational", "New England - commercial"),
        col = c(4, 4, 2, 2, 3, 3, 1, 1), lty = c(1, 3, 1, 3, 1, 3, 1, 3), lwd = 2,
        bty = "n", y.intersp = 0.9, cex = 0.85)

UScatch <- rowSums(datnew, na.rm = T)
write.csv(data.frame(UScatch), file = "data/total_mainUS_catches.csv", row.names = T)

```

```{r}
#| echo: false

```
