# NOAA Fisheries recreational landings data

Here we will partition recreational landings data from NOAA Fisheries’ Marine Recreational Information Program (MRIP), a standardized, peer-reviewed suite of recreational fishing surveys used to produce catch and effort estimates for quota monitoring and stock assessment in the United States. MRIP is a standardized survey suite that employs a complex statistical sampling design stratified across time, regions of the coast, and different fishing modes, which allows the intercept data to be scaled up by effort estimates to calculate total landings.

## Data access and upload

Data are downloaded directly from the Miami server using the files provided to SERO for ACL monitoring. The current location is: M://SFD/SECM-SFD/ACL/2025_Aug19_MRIP_FES/. The landings are in FES MRIP and the units are in pounds whole weight.

```{r}
#| warning: false
#| echo: true

# clear workspace
rm(list = ls())

# read in data file
load("data/mrip_fes_rec81_25wv2_23June25.RData")  # most recent file on SFD server
head(dat)

#apply(dat[2:18], 2, table, useNA = "always")

# confirm nomenclature for dolphin and subset by species --------------------------
table(dat$NEW_COM[grep("dolphin", dat$NEW_COM)])
table(dat$NEW_SCI[grep("dolphin", dat$NEW_COM)])

d <- dat[which(dat$NEW_COM == "dolphin"), ]
#apply(d[2:18], 2, table, useNA = "always")
d <- d[which(d$YEAR < 2025), ]    # take out 2025 because incomplete year      

```

## Separate Gulf and Atlantic recreational landings

Now we filter the data set to separate out the Gulf versus Atlantic landings.

```{r}
#| echo: true

# codes for WFL regions: 1 (NW FL Panhandle- Escambia to Dixie co.), 2 (SW FL Peninsula- Levy to Collier co), 3 (FL Keys- Monroe co.)
table(d$FL_REG, useNA = "always")   

# subregional codes: 6 is Atlantic and 7 is Gulf
table(d$FL_REG, d$SUB_REG, useNA = "always")
unique(d$NEW_STA)

lis <- c("TX", "LA", "LA/MS", "MS", "AL", "AL/FLW")  # filter out the Gulf states
d1 <- d[-which(d$NEW_STA %in% lis), ]
unique(d1$NEW_STA)

table(d1$FL_REG, d1$NEW_STA, useNA = "always")  

dim(d1)
d1 <- d1[-which(d1$FL_REG == 1 | d1$FL_REG == 2), ]   # filter out FL regions 1 and 2 but retain 3 (FL Keys)
dim(d1)
d1 <- d1[-which(is.na(d1$FL_REG) & d1$NEW_STA == "FLW"), ]  # filter out FLW with unknown FL region
dim(d1)

table(d1$FL_REG, d1$NEW_STA, useNA = "always")  # retained are all Atlantic states plus region 3 of Florida
```

Now we can summarize the total dolphin landings in weight for the U.S. Atlantic, and compare to previous values sent for quota monitoring. We see that the values are almost identical.

```{r}
#| echo: true

# summarize by year and compare --------------------------

tot1 <- tapply(d1$LBSEST_SECWWT, d1$YEAR, sum, na.rm = T)
head(tot1)

old <- read.csv("data/RecreationalDolphinLandings_SAandGOM_MLarkin.csv", skip = 6)
plot(old$Year, old$ATL, type = "l", xlab = "year", ylab = "total catch (pounds)", lwd = 4, 
     xlim = c(1980, 2025), ylim = c(7*10^6, 3.3*10^7))
lines(as.numeric(names(tot1)), tot1, lwd = 2, col = 2, lty = 2)
legend("topright", c("previous data", "new data"), col = c(1, 2), lwd = c(4, 2), lty = c(1, 2))

```

## Partition catches by area and sector

Now we will partition out the total catches according to the dolphin MSE areas. The areas are as follows:

-   FLK: Florida Keys to Indian River County County (FL)
-   NCFL: Brevard (FL) to Southern NC (South of Hatteras)
-   NNC: Northern NC (North of Hatteras to NC/VA border)
-   VBM: Virginia to Maine

```{r}
#| echo: true

# separate out 4 regions --------------------------

d1$region <- "VBM"

# select for FLK
# FL_REG codes: 4: SE FL- Miami-Dade to Indian River County.; 5: NE FL- Brevard to Nassau County
table(d1$FL_REG, d1$NEW_STA, useNA = "always")
d1$region[which(d1$FL_REG == 3 | d1$FL_REG == 4)] <- "FLK"

# select for NCFL
d1$region[which(d1$FL_REG == 5)] <- "NCFL"
lis <- c("SC", "GA", "FLE/GA")           # NCFL states
d1$region[which(d1$NEW_STA %in% lis)] <- "NCFL"
# NC_REG codes: N: North of Cape Hatteras; S: South of Cape Hatteras
table(d1$NC_REG, d1$NEW_STA, useNA = "always")
d1$region[which(d1$NC == "S")] <- "NCFL"

# select for NNC
d1$region[which(d1$NC == "N")] <- "NNC"

table(d1$FL_REG, d1$NEW_STA, d1$region)
table(d1$NC_REG, d1$NEW_STA, d1$region)  # check classification

```

Now we will separate out the catches by recreational sector and plot the totals by region and sector.

```{r}
#| warning: false
#| echo: true

# separate out by for hire vs. private rec --------------------------
d1$sector <- "ForHire"

d1$sector[which(d1$NEW_MODEN == "Shore")] <- "Rec"
d1$sector[which(d1$NEW_MODEN == "Priv")]  <- "Rec"

table(d1$sector, d1$NEW_MODEN, useNA = "always")    # check classification

tot <- tapply(d1$LBSEST_SECWWT, list(d1$YEAR, d1$region, d1$sector), sum, na.rm = T)

head(tot)

fin <- data.frame(cbind(tot[, , 1], tot[, , 2]))
names(fin) <- c("Hire_FLK", "Hire_NCFL", "Hire_NNC", "Hire_VBM", "Rec_FLK", "Rec_NCFL", "Rec_NNC", "Rec_VBM")
head(fin)

matplot(rownames(fin), fin, lty = c(rep(2, 4), rep(1, 4)), col = rep(1:4, 2), type = "l", lwd = 2, 
        xlab = "year", ylab = "total catch (pounds)")
legend("topright", colnames(fin), lty = c(rep(2, 4), rep(1, 4)), col = rep(1:4, 2), lwd = 2, ncol = 2, bty = "n")

# check that totals match
cor(rowSums(fin, na.rm = T), tot1)
mean(rowSums(fin, na.rm = T) - tot1)   # very tiny differences

```

Here are the catches (in pounds) for years 1981 to 2024.

```{r}
#| warning: false
#| echo: true

fin
```

## Looking at extent of discarding

Discard estimates (known as B2 estimates) are provided in numbers by the MRIP survey but are not available in the Southeast Region Headboat Survey. We can look at the overall numbers of dolphinfish caught versus reported released from the MRIP survey to get an understanding of the extent of discarding.

```{r}
#| warning: false
#| echo: true
#| fig-width: 10

d1$perdis <- d1$B2 / (d1$AB1 + d1$B2)  # percent discarded is B2 (fish released alive) / total

# calculate mean discard rate by year, region, sector ----------------
dis <- tapply(d1$perdis, list(d1$YEAR, d1$region, d1$sector), mean, na.rm = T)
disr <- tapply(d1$perdis, list(d1$region, d1$sector), mean, na.rm = T)

matplot(rownames(dis), cbind(dis[, , 1], dis[, , 2]), lty = c(rep(2, 4), rep(1, 4)), col = rep(1:4, 2), 
        type = "l", lwd = 2, xlab = "year", ylab = "% dolphin discarded", ylim = c(0, 0.5))
legend("top", c(paste("For hire - ", colnames(dis)), paste("Private - ", colnames(dis))), 
       lty = c(rep(2, 4), rep(1, 4)), col = rep(1:4, 2), lwd = 2, ncol = 2, bty = "n")

round(disr * 100, 2)  # discarding rates by region and sector

```

According to the MRIP survey and the reported numbers, discarding generally increases over time, excluding some highly anomalous estimates from the early years which are driven by low sample sizes. Discarding is generally higher in the private sector compared to the for-hire sector, and is highest in the Florida Keys region.

## Split the catches by season

The dolphin MSE operates on a seasonal time step and thus the catches and discards must by parsed apart by individual season and year. This is not straightforward because MRIP estimates are reported in two-month waves (Jan - Feb, Mar - Apr, May - Jun, etc.), and the seasons are 3-month periods (Dec - Feb, Mar - May, Jun - Aug, Sep - Nov). For the two waves that have to be split into seasons (e.g., December is lumped with January and February; June is lumped with July and August), we split the wave 50/50%, i.e, we make the assumption that half of the catch in that wave occurs in one season and the other half occurs in the next season.

```{r}
#| warning: false
#| echo: true

# define seasons -----------------------------
table(d1$WAVE, useNA = "always")

d1$YRWAVE <- paste0(d1$YEAR, "_", d1$WAVE)

totw <- tapply(d1$LBSEST_SECWWT, list(d1$YRWAVE, d1$region, d1$sector), sum, na.rm = T)

f <- data.frame(cbind(totw[, , 1], totw[, , 2]))
names(f) <- c("Hire_FLK", "Hire_NCFL", "Hire_NNC", "Hire_VBM", "Rec_FLK", "Rec_NCFL", "Rec_NNC", "Rec_VBM")
head(f)

f$YEAR <- as.numeric(substr(rownames(f), 1, 4))
f$WAVE <- as.numeric(substr(rownames(f), 6, 8))
head(f)   # table with catches separated by wave and year
f[is.na(f)] <- 0

f <- f[which(f$YEAR >= 1985), ]
f <- f[which(f$YEAR <= 2024), ]

table(f$YEAR)
table(f$WAVE)  

# deal with NAs in many locations ---------------
# calculate percentages of catch by wave, for each region and fleet
avw <- tapply(d1$LBSEST_SECWWT, list(d1$WAVE, d1$region, d1$sector), sum, na.rm = T)
avw <- avw[-1, , ]
avw <- data.frame(cbind(avw[, , 1], avw[, , 2]))
names(avw) <- c("Hire_FLK", "Hire_NCFL", "Hire_NNC", "Hire_VBM", "Rec_FLK", "Rec_NCFL", "Rec_NNC", "Rec_VBM")
avwp <- apply(avw, 2, function (x) x / sum(x, na.rm = T)) 
#colSums(avwp)

# go through each column, find NAs and fill in waves by percentages
pre_colsum <- colSums(f, na.rm = T)  # column sums to check later

lis <- which(f$WAVE == 0)  # list of rows with NA wave data

for (i in 1:8) {      # loop through each column
  for(j in lis) {     # loop through list of NA waves
    if (f[j, i] > 0) {  # if there are NA wave catches, split by known proportion from other years
      miscatch <- avwp[, i] * f[j, i]   # missing catch is known proportion * NA wave catch
      f[which(f$YEAR == f$YEAR[j] & f$WAVE != 0), i] <- f[which(f$YEAR == f$YEAR[j] & f$WAVE != 0), i] + miscatch  # add in missing catch
      f[j, i] <- 0   # set to zero after catch is attributed to waves
    }
  }
}
f[lis, 1:8] # should be all zeros now
pos_colsum <- colSums(f, na.rm = T) 
pre_colsum[1:8] == pos_colsum[1:8]  # check that catch was parsed correctly - column sums should be identical 

f <- f[which(f$WAVE != 0), ]  # remove NA waves which now contain no catch

# split waves 3 and 6 50/50% to parse into seasons ------------------
lis <- which(f$WAVE == 3 | f$WAVE == 6)  # list of 3 and 6 waves for all years

for (i in lis) {          # go through list of waves
  temp <- f[i, ] / 2      # split catch in half
  temp[9] <- f$YEAR[i]    # reformat year and wave data columns
  temp[10] <- f$WAVE[i]
  temp <- rbind(temp, temp)   # create two half-waves and relabel with new wave reference number
  temp[1, 10] <- temp[1, 10] - 0.5
  temp[2, 10] <- temp[2, 10] + 0.5
  f <- rbind(f, temp)         # concatenate with main data table
  }

f <- f[-lis, ]  # remove waves that have been split in half

pos_colsum <- colSums(f, na.rm = T) 
pre_colsum[1:8] == pos_colsum[1:8]  # check that catch was parsed correctly - column sums should be identical 

f <- f[order(f$YEAR, f$WAVE), ]  # reorder table in chronological order

f$YEAR[which(f$WAVE == 6.5)] <- f$YEAR[which(f$WAVE == 6.5)] + 1  # add year to December wave so it gets summed with following year
f$WAVE[which(f$WAVE == 6.5)] <- 0.5  # change December wave to small number so it gets summed with Jan/Feb

table(f$WAVE)
f$seas <- cut(f$WAVE, breaks = c(0, 1.5, 3, 4.5, 6))
#table(f$seas)
f$seas1 <- NA
f$seas1[which(f$seas == "(0,1.5]")] <- "DJF"  # Dec Jan Feb is 0.5 and 1
f$seas1[which(f$seas == "(1.5,3]")] <- "MAM"  # Mar Apr May is 2 and 2.5
f$seas1[which(f$seas == "(3,4.5]")] <- "JJA"  # Jun Jul Aug is 3.5 and 4
f$seas1[which(f$seas == "(4.5,6]")] <- "SON"  # Sep Oct Nov is 5 and 5.5

temp <- tapply(f[, 1], list(f$seas1, f$YEAR), sum, na.rm = T)
temp1 <- matrix(temp)
for (i in 2:8) {
  temp2 <- matrix(tapply(f[, i], list(f$seas1, f$YEAR), sum, na.rm = T))
  temp1 <- cbind(temp1, temp2)
  }

findat <- data.frame(sort(rep(as.numeric(colnames(temp)), 4)), 
                     rep(rownames(temp), ncol(temp)), temp1)
names(findat) <- c("year", "season", names(f)[1:8])

colSums(f[, 1:8], na.rm = T) == colSums(findat[, 3:10], na.rm = T)  # check that total catch is equal before and after parsing
```

We plot the results by region and sector, over the time series of year and season. We can clearly see the seasonality in the catch by region.

```{r}
#| echo: true
#| fig-width: 10

findat$yrseas <- findat$year + as.numeric(as.factor(findat$season))/4-0.125
matplot(findat$yrseas, findat[, 3:10]/10^6, lty = c(rep(2, 4), rep(1, 4)), col = rep(1:4, 2), type = "l", lwd = 2, 
        xlab = "year", ylab = "total catch (millions of pounds)")
legend("topright", colnames(findat[, 3:10]), lty = c(rep(2, 4), rep(1, 4)), col = rep(1:4, 2), lwd = 2, ncol = 2, bty = "n")

# check that totals match
cbind(colSums(findat[, 3:10], na.rm = T), colSums(fin[-c(1:5), ], na.rm = T))
# the values should be close but slightly >1 because the last data set includes December 1985 data as part of 1986
colSums(findat[, 3:10], na.rm = T) / colSums(fin[-c(1:5), ], na.rm = T)
```
