# Analysis of international landings in the high seas

Here we will explore commercial and recreational landings data as estimated by the [Sea Around Us Project](https://www.seaaroundus.org/data/#/eez). This data set includes reported data for fisheries worldwide as well as reconstructed estimates for unreported landings. For each country, project scientists assemble available data on landings and use additional sociological and fishing data plus expert information and opinion, to produce their best estimates of total landings from all fishing sectors and for all species globally.

## Data access and upload

The high seas data are accessed in the Basic Search [High seas section of the site](https://www.seaaroundus.org/data/#/highseas). The individual regions can be clicked on (e.g., Atlantic, Western Central) and then the data are downloaded using the Download Data button.  

The current version of data in use is version 50.1, accessed June 2024 (the data files can be found on github in the SEFSC-dolphin-analyses/data /SAU/ folder). This was still the most recent version of data available as of June 2025.

## Read in the high seas landings

First we read in the high seas landings for the Western Central Atlantic and Northwest Atlantic zones. These are equivalent to FAO areas 21 and 31.

```{r}
#| warning: false
#| echo: true

# clear workspace
rm(list = ls())

if(!require("pals")) install.packages("pals")
library(pals)

# find data files
hslis <- dir("data/SAU/")[grep("HighSeas", dir("data/SAU/"))]

# identify areas 21 and 31
hslis[1]; hslis[3]   

#download data
h1 <- read.csv(paste0("data/SAU/", hslis[1]))   
h3 <- read.csv(paste0("data/SAU/", hslis[3]))

# confirm correct areas
#unique(h1$area_name)
#unique(h3$area_name)

hs <- rbind(h1, h3) 
```

We extract the data for dolphinfish and look at the characteristics of the landings. Most of the landings are in the Western Central Atlantic, and the primary fishing entities are Venezuela and Saint Lucia. The fishing sector is all industrial, and the vast majority of the catch is reported landings, with only a tiny fraction of discards or unreported landings listed in the database. The gear type is primarily long line.

```{r}
#| warning: false
#| echo: true

# check species name and filter out dolphinfish
unique(hs$common_name)[grep("dolphin", unique(hs$common_name))]
hs <- hs[which(hs$common_name == "Common dolphinfish"), ]

# look at the categories within each column of data
apply(hs[1:13], 2, table)

# convert from metric tonnes to pounds
hs$pounds <- hs$tonnes * 1000 * 2.20462

# produce barplot of key data fields
par(mar = c(3, 20, 3, 1), mfrow = c(3, 2), mex = 0.5)
barplot(tapply(hs$pounds/10^6, hs$area_name, sum, na.rm = T), las = 2, xlab = "millions of pounds", main = "Area name", horiz = T)
#barplot(tapply(hs$pounds/10^6, hs$area_type, sum, na.rm = T), las = 2, ylab = "millions of pounds", main = "Area type")
#barplot(tapply(hs$pounds/10^6, hs$year, sum, na.rm = T), las = 2, ylab = "millions of pounds", main = "Year")
barplot(tapply(hs$pounds/10^6, hs$fishing_entity, sum, na.rm = T), las = 2, xlab = "millions of pounds", main = "Fishing entity", horiz = T)
barplot(tapply(hs$pounds/10^6, hs$fishing_sector, sum, na.rm = T), las = 2, xlab = "millions of pounds", main = "Fishing sector", horiz = T)
barplot(tapply(hs$pounds/10^6, hs$catch_type, sum, na.rm = T), las = 2, xlab = "millions of pounds", main = "Catch type", horiz = T)
barplot(tapply(hs$pounds/10^6, hs$reporting_status, sum, na.rm = T), las = 2, xlab = "millions of pounds", main = "Reporting status", horiz = T)
barplot(tapply(hs$pounds/10^6, hs$gear_type, sum, na.rm = T), las = 2, xlab = "millions of pounds", main = "Gear type", horiz = T)

round(tapply(hs$pounds, hs$catch_type, sum, na.rm = T) / sum(hs$pounds) * 100, 2)   # << 1% are discards
round(tapply(hs$pounds, hs$reporting_status, sum, na.rm = T) / sum(hs$pounds) * 100, 2)  # << 1% are unreported

```

## Summarize the high seas landings

We plot the annual catches by fishing entity to look at trends over time.

```{r}
#| warning: false
#| echo: true

hs$fishing_entity <- as.factor(hs$fishing_entity)
hs$fishing_entity <- droplevels(hs$fishing_entity)

par(mar = c(5, 5, 3, 1), mfrow = c(1, 1))
tab <- tapply(hs$pounds, list(hs$year, hs$fishing_entity), sum, na.rm = T)
matplot(as.numeric(rownames(tab)), tab/10^6, type = "l", col = glasbey(ncol(tab)), lty = 1, lwd = 2, xlab = "year", ylab = "total catch (millions of pounds)", 
        main = "High seas Western Central Atlantic dolphin catch")
legend("topleft", colnames(tab), col = glasbey(ncol(tab)), lty = 1, lwd = 2, cex = 0.8, bty = "n")
```

Finally we output the subset of high seas dolphin catch for the two areas. Because we have already accounted for the USA landings in our national reporting and are aiming to only summarize international landings here, we filter out the USA landings from the Sea Around Us database.

```{r}
#| warning: false
#| echo: true

hs <- hs[which(hs$fishing_entity != "USA"), ]

hs$year <- factor(hs$year, levels = 1950:2019)
tab <- tapply(hs$pounds, list(hs$year, hs$area_name), sum, na.rm = T)
tab[is.na(tab)] <- 0
tail(tab, 40)

# write summary to merge with EEZ data 
write.csv(tab, file = "data/SAU/high_seas_by_year.csv")
```

## Distribute the high seas landings across quarters of the year

The international landings are only reported as annual catches and do not have any month or season associated with them. However, the MSE operating model requires seasonal catches. The majority of the international landings are caught by longline, so we will assume that the fleet distribution of the U.S. pelagic longline is representative of the seasonality of the international high seas fleets. Now we parse the annual catches according to the seasonality of the U.S. fleet.

At the time of analysis, the SAU database only contained catches to 2019, so we interpolated the 2019 values to cover years 2020 - 2022.

```{r}
#| echo: true

# read in the percentage of the catch by area and year-quarter combinations from the U.S. PLL
load("data/outputs/per_PLLcatch_by_area_yearquarter.RData")  

tab <- tab[which(rownames(tab) >= 1997), ]  # PLL data are only available from 1997 on 
#tab[nrow(tab), ]
tab <- rbind(tab, tab[nrow(tab), ], tab[nrow(tab), ], tab[nrow(tab), ])  # interpolate 2019 catches for 2020-2022
tail(tab)

NED <- percatch[, , 1]   # create matrices for final catch data
NCA <- percatch[, , 1]

for (i in 1:ncol(percatch))  { 
  NED[, i] <- percatch[, i, 7] * tab[i, 1]   # these are the percentages for the NED (Atlantic Northwest)
  NCA[, i] <- percatch[, i, 1] * tab[i, 2]   # these are the percentages for the NCA (Western Central Atlantic)
  }

# plot the parsed out data by year-quarter
yrs <- sort(rep(as.numeric(colnames(NCA)), 4))
qrt <- rep(1:4, length(yrs)/4)

matplot(yrs + qrt/4, cbind(matrix(NED), matrix(NCA))/10^6, lty = 1, lwd = 2, 
        col = 2:3, type = "l", main = "International high seas catches by quarter", 
        xlab = "", ylab = "total catch (millions of pounds)")
legend("topleft", c("NED", "NCA"), lwd = 2, col = 2:3, bty = "n")

# save NED data in this format because it needs to be combined with Canada EEZ catches
save(NED, file = "data/outputs/NED_year_quarter.RData")

```

Finally, we format the data in the format for input into the operating model. We only include NCA here, as the NED high seas catches must be combined with international catches in territorial waters of Canada and France in the NED region.   

```{r}
#| echo: true

# format for operating model
flt <- rep("Intl", length(yrs))
are <- rep("NCA", length(yrs))
findat <- data.frame(cbind(yrs, qrt, flt, are, matrix(NCA)))
names(findat) <- c("Year", "Quarter", "Fleet", "Area", "Catch_lbs")
findat$Catch_lbs <- as.numeric(findat$Catch_lbs)
head(findat)

#plot(findat$Catch_lbs, type = "l")
#plot(findat$Catch_lbs ~ factor(findat$Area))
#plot(findat$Catch_lbs ~ factor(findat$Quarter))
#plot(findat$Catch_lbs ~ factor(findat$Year))
#apply(findat, 2, table)
#barplot(tapply(findat$Catch_lbs, list(findat$Quarter, findat$Area), sum, na.rm = T), beside = T, col = rainbow(4))

write.csv(findat, file = "data/FINAL_files/NCA_Intl_TomFormat.csv", row.names = FALSE)
```
